Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-10 treestats_paper.tex
O() fixup: closes #12
added section on restrictions/properties: closes #17

@@

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$. Therefore, the overall running time
of Algorithms~\algref{B} is
\[
O(n + \rho \ln n) \log_2 n
\]
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root once. Similarly,
edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

>>>>

%%%%%%%%%%%%%%%%%%
\subsection*{General properties}

The statistics defined here are all \emph{additive} along the genome:
if we have computed a statistic in equal-size windows along a chromsome,
then the value of the statistic for the entire chromosome is equal to
the average of the values in those windows.
Some commonly-used statistics (e.g., $F_{ST}$ or Tajima's $D$)
are not additive, but are ratios of additive statistics, so can be easily computed in this way.
Multilocus statistics that measure linkage disequilibrium or haplotype lengths
do not fall in this category.
We have also defined site statistics to be additive across \emph{alleles},
to allow the definitions to apply to sites with more than two alleles.
This decision has some nice properties, but is not the only reasonable choice.

We also require that site statistics be computable from the genotype matrix
(i.e., without knowledge of the trees),
and, if the statistic is polarised, knowledge of the ancestral allele.
We often would also like our statistics to be unaffected by the addition of monomorphic sites,
which are usually omitted from genetic variation data.
To ensure this we have required that the summary function returns zero when give both zero weight
and the total weight: $f(0) = f(\tiw) = 0$.
This also makes branch and node statistics unchanged by portions of the tree that do not lie on a path
between any of the samples that have nonzero weight,
so that in particular,
\emph{simplification} of the tree sequence will not change the value of the statistic.
However, in practice this requirement can be relaxed.

@@

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$. Therefore, the overall running time
of Algorithms~\algref{B} is of order $(n = \rho \ln n)\log_2 n$, which is
\[
    O\left(n \log n + \rho \log^2 n \right)
\]
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root once. Similarly,
edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.
